A Kochen-Specker system has at least 21 vectors\\ {\small (extended abstract)}

### 2014-09-11 paper/main.tex
paper: add non-edge-triconnected graph figure
Improved connectivity proofs

\usepackage{algorithm2e}
\usepackage{breakurl}
\usepackage{hyperref}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{wrapfig}
\usepackage{amsthm}
\usepackage{tikz}

@@

\newcounter{main}
\newtheorem{prop}[main]{Proposition}
\newtheorem{comp}[main]{Computation}
\newtheorem{cor}[main]{Corollary}
\newtheorem{thm}[main]{Theorem}
\newtheorem{lem}[main]{Lemma}
\newtheorem{fact}[main]{Fact}
\theoremstyle{definition}
\newtheorem{dfn}[main]{Definition}
\newtheorem*{spin}{SPIN Axiom \cite{ck09}}
\theoremstyle{remark}
\newtheorem{rem}[main]{Remark}

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlapping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is edge-biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not edge-biconnected graph~$G$.
Then we can decompose
this graph into its edge-biconnected components.  A edge-biconnected components
is a maximal subgraph that is edge-biconnected.
We can consider the graph of edge-biconnected components~$B(G)$:
two edge-biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of edge-biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the edge-biconnected components in the graph,
would be itself edge-biconnected.  Thus~$B(G)$ is a tree.

Furthermore suppose, reasoning towards contradication,
that~$G$ is a minimal Kochen-Specker graph.
Now consider a leaf~$A$ in the edge-biconnected component graph~$B(G)$.
That is, $A$ is a edge-biconnected component that is connected to only
one other edge-biconnected component.
Let~$B$ be the remainder of the graph.
There is exactly one pair~$(a,b)$ with~$a$ in~$A$ and~$b$ in~$B$
and~$a$ adjacent to~$b$.

Note that~$A$ and~$B$ are~$010$-colorable.
Suppose that~$a$ does not have fixed color~$1$ in~$A$.
Then there is a~$010$-coloring of~$A$
which assigns~$0$ to~$a$.
But then~$G$ is~$010$-colorable, quod non.
Thus~$A$ is an embeddable graph with a vertex with fixed color~$1$.
Similarly~$B$ is an embeddable graph with a vertex with fixed color~$1$.

Let~$A'$ denote the graph containing two copies of~$A$
connected at the copies of the vertex with fixed color~$1$.
Using the reasoning in the proof of Lemma~\ref{lem:biconn},
we see~$A'$ is a Kochen-Specker graph that is not edge-biconnected.
Define~$B'$ similarly. Observe that~$A'$ and~$B'$ must have the same
number of vertices as~$G$ by the minimality of~$G$.

The graph~$A'$ has minimal vertex-order~$3$.
Thus every vertex of~$A$ has minimal order~$3$ except for~$a$,
which has minimal order~$2$.
Suppose~$a$ has minimal order~$3$ in~$A$.
Then by Computation~\ref{comp:bic1},
the graph~$A$ has at least~$16$ vertices.
Thus~$A'$ has at least~$32$ vertices, quod non.

Thus~$a$ must have order~$2$ in~$A$.
Let~$c,d \in A$ be the vertices connected to~$a$ in~$A$.
For~$a$ to have fixed color~$1$ in~$A$,
the vertices~$c$ and~$d$ must have fixed color~$0$ in~$A$.
By Computation~\ref{comp:bic2},
the graph~$A$ has at least~$15+1$ vertices.
Thus~$A'$ has at least~$32$ vertices. Contradiction.
\end{proof}

We can go one step further.
\begin{prop}
    A minimal Kochen-Specker graph is edge-triconnected.
    That is: removing any two edges keeps the graph connected.
\end{prop}
\begin{proof}
    Given a minimal Kochen-Specker graph~$G$.
    Suppose it is not edge-triconnected.
    Then it splits into two graphs~$A$ and~$B$
    together with verteces~$a_1, a_2 \in A$
    and~$b_1,b_2 \in B$
    such that~$(a_1, b_1)$ and~$(a_2,b_2)$
    are the only edges between~$A$ and~$B$.
    Note that~$A$ and~$B$ must be $010$-colorable,
    for otherwise~$G$ would not be a minimal Kochen-Specker graph.
    \begin{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 = b_2$.
                Then~$G$ is not edge-biconnected.  Contradiction with
                    Proposition~\ref{prop:ks-biconn}.
    \item Suppose~$a_1 \neq a_2$ and~$b_1 = b_2$.
        \label{tricase1}
        Suppose~$b_1=b_2$ does not have a fixed color in~$B$.
        Then any coloring of~$A$ can be extended with some coloring
        in~$B$ to a coloring of~$G$.  Contradiction.
        Apparently~$b_1=b_2$ has a fixed color in~$B$.
        \begin{enumerate}
            \item
            Suppose~$b_1=b_2$ has fixed color~$1$ in~$B$.
            Note~$\#B \geq 16$.

@@

            Thus the type of~$(a_1,a_2)$ in~$A$ cannot contain~$(0,0)$.
            Thus~$\#A \geq 16$.
            Consequently~$\#G \geq 32$. Contradiction with minimality.

            \item
            Apparently~$b_1=b_2$ has fixed color~$0$ in~$B$.
            Note~$\#B \geq 15$.

@@

            The type of~$(a_1,a_2)$ in~$A$ cannot contain~$(1,0)$
            or~$(0,1)$ for otherwise~$G$ can be colored.
            It also cannot contain~$(1,1)$ as~$a_1$ and~$a_2$
            are adjacent.
            Thus both~$a_1$ and~$a_2$ have fixed color~$0$ in~$A$.
            Hence~$\#A\geq 17$. (?) % FIXME check and ref
            Consequently~$\#G \geq 32$. Contradiction
            with minimality.
        \end{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 \neq b_2$.
            This leads to a contradication in the same way
            as in case~\ref{tricase1}.
    \item Apparently~$a_1 \neq a_2$ and~$b_1 \neq b_2$.
        The type of~$(a_1,a_2)$ in~$A$ cannot contain~$(0,0)$,
        for otherwise~$G$ is colorable.
        Similarly, the type of~$(b_1,b_2)$ in~$B$ cannot contain~$(0,0)$.
        Thus both~$\#A \geq 16$ and~$\#B \geq 16$.
        Hence~$\#G \geq 32$. Contradiction with minimality. \qedhere
    \end{enumerate}

\end{proof}

>>>>

\usepackage{algorithm2e}
\usepackage{breakurl}
\usepackage{hyperref}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{mdwlist}
\usepackage{amsmath}
\usepackage{wrapfig}
\usepackage{picins}
\usepackage{amsthm}
\usepackage{tikz}

@@

\newcounter{main}
\newtheorem{prop}[main]{Proposition}
\newtheorem{comp}[main]{Computation}
\newtheorem{cor}[main]{Corollary}
\newtheorem{thm}[main]{Theorem}
\newtheorem{lem}[main]{Lemma}
\newtheorem{fact}[main]{Fact}
\theoremstyle{definition}
\newtheorem{dfn}[main]{Definition}
\newtheorem{exa}[main]{Example}
\newtheorem*{spin}{SPIN Axiom \cite{ck09}}
\theoremstyle{remark}
\newtheorem{rem}[main]{Remark}

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlapping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.

\begin{comp}\label{comp:bic1}
    There are no embeddable graphs with fixed color~$1$
    on less than 17 vertices.
\end{comp}

We are ready to prove that a minimal Kochen-Specker graph
is edge-biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Given a minimal Kochen-Specker graph~$G$.
\parpic(0cm,1cm)[r]{%
    \begin{tikzpicture}[thick,scale=1.4,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw [rotate=90] (0,-0.8) ellipse (0.6 and 0.5) ;
            \draw [rotate=90] (0,0.8) circle (0.6 and 0.5) ;
            \draw 
                  (-0.7, 0.1) node[left] {$a$}
                  (0.7, 0.1) node[right] {$b$}
                  (-0.7, 0.1) node[p] {} -- (0.7, 0.1) node[p] {}
                  (-0.8, -0.6) node[below] {$A$}
                  (0.8, -0.6) node[below] {$B$}
                ;
        \end{tikzpicture}}
Recall it must be connected.  Suppose it is not edge-biconnected.
Then there must be an edge~$(a,b)$ in~$G$,
which removal disconnects~$G$.
Thus~$G$ decomposes into two connected graphs~$A$ and~$B$
such that~$a\in A$, $b \in B$ and~$(a,b)$ is the only edge
between~$A$ and~$B$.
Clearly~$A$ and~$B$ are embeddable.

Note that~$A$ must be~$010$-colorable,
for if it were not~$010$-colorable,
then~$A$ is a Kochen-Specker graph,
in contradiction with~$G$'s minimality.
Similarly~$B$ is~$010$-colorable.
Suppose there is a~$010$-coloring of~$A$ in which~$a$ is colored~$0$.
Then we can extend this coloring with any $010$-coloring of~$B$
to a~$010$-coloring of~$G$, which is absurd.
Thus~$a$ must have fixed color~$1$ in~$A$.
Similarly~$b$ must have fixed color~$1$ in~$B$.
Thus by Computation~\ref{comp:bic1},
we have~$\#A \geq 17$ and~$\#B \geq 17$.
Consequently~$\#G \geq 34$.
Contradiction with~$G$'s minimality.
\end{proof}

We can go one step further.

\begin{prop}\label{prop:ks-triconn}
    A minimal Kochen-Specker graph is edge-triconnected.
    That is: removing any two edges keeps the graph connected.
\end{prop}

Again, we need some preparation for the proof.  First, we generalize
the notion of fixed color.

\begin{dfn}
   Given a graph~$G$ together with selected vertices~$v_1, \ldots, v_n \in G$.
   Let~$C(G)$ denote the set of $010$-colorings of~$G$.
   The \keyword{type of~$(v_1,\ldots,v_n)$ (in~$G$)}
   is the set of all possible ways it can be colored;
   that is: $t_G(v_1, \ldots, v_n) := \{ (c(v_1), \ldots, c(v_n));
                \ c \in C(G) \}.$ 
    A type of~$n$ vertices is
    called an~\keyword{$n$-type}.
\end{dfn}

Note that~$v$ has the~$1$-type~$(1)$ in~$G$ if and only if
it has fixed color~$1$ in~$G$.
Just like points with fixed color,
we are interested in types,
because they do not occur in small graphs.

\begin{comp}\label{comp:types}
    We have enumerated all embeddable graphs of less than~$17$ vertices
    and determined a lower bound at which a particular $1$- or $2$-type occurs.
    \begin{center}
    \begin{tabular}{ll}
        $1/2$-type & $\#G$ \\ \hline
        $\{(0,0), (1,0), (0,1)\}$ &
            non-trivially $\geq 10$\\ 
        $\{(0,0), (1,0), (1,1)\}$ & $\geq 10$ \\
        $\{(0,0), (0,1), (1,1)\}$ & $\geq 10$ \\
        $\{(0,0), (0,1))\}$ & $\geq 15$ \\
        $\{(0,0), (1,0))\}$ & $\geq 15$ \\
        $\{(0)\}$ & $\geq 15$ \\
        $\{(0,1),(1,0)\}$ & $\geq 16$ \\
        other & $\geq 17$
    \end{tabular}
    \end{center}
    The type~$\{(0,0),(1,0),(0,1)\}$ occurs in the embeddable two-vertex
    graph.  However, since
    \begin{tikzpicture}[thick,scale=1.4,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw
            (0,0) node[p]{} -- (0.3,0) node[p]{} ;
    \end{tikzpicture}
    This occurance is called trivial, for the two vertices are adjacent.
\end{comp}

\begin{proof}[Proof of Proposition~\ref{prop:ks-triconn}]

    Given a minimal Kochen-Specker graph~$G$.

\parpic(0cm,1cm)[r]{%
\begin{tikzpicture}[thick,scale=1.4,
p/.style={circle, draw, fill=white,
                inner sep=0pt, minimum width=5pt}]
    \draw [rotate=90] (0,-0.8) ellipse (0.8 and 0.5) ;
    \draw [rotate=90] (0,0.8) circle (0.8 and 0.5) ;
    \draw 
          (-0.7, 0.3) node[left] {$a_1$}
          (-0.7, -0.3) node[left] {$a_2$}
          (0.7, 0.3) node[right] {$b_1$}
          (0.7, -0.3) node[right] {$b_2$}
          (-0.7, -0.3) node[p] {} -- (0.7, -0.3) node[p] {}
          (-0.7, 0.3) node[p] {} -- (0.7, 0.3) node[p] {}
          (-0.8, -0.8) node[below] {$A$}
          (0.8, -0.8) node[below] {$B$}
        ;
\end{tikzpicture}}

    Suppose it is not edge-triconnected.
    Then it splits into two graphs~$A$ and~$B$
    together with verteces~$a_1, a_2 \in A$
    and~$b_1,b_2 \in B$
    such that~$(a_1, b_1)$ and~$(a_2,b_2)$
    are the only edges between~$A$ and~$B$.
    Note that~$A$ and~$B$ must be $010$-colorable,
    for otherwise~$G$ would not be a minimal Kochen-Specker graph.

    \begin{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 = b_2$.
                Then~$G$ is not edge-biconnected.  Contradiction with
                    Proposition~\ref{prop:ks-biconn}.
    \item Suppose~$a_1 \neq a_2$ and~$b_1 = b_2$.
        \label{tricase1}
        Suppose~$b_1=b_2$ does not have a fixed color in~$B$.
        Then any coloring of~$A$ can be extended with some coloring
        in~$B$ to a coloring of~$G$.  Contradiction.
        Apparently~$b_1=b_2$ has a fixed color in~$B$.
        \begin{enumerate}
            \item
            Suppose~$b_1=b_2$ has fixed color~$1$ in~$B$.
            Note~$\#B \geq 17$ by Computation~\ref{comp:types}.

@@

            Thus the type of~$(a_1,a_2)$ in~$A$ cannot contain~$(0,0)$.
            Thus, by Computation~\ref{comp:types}, $\#A \geq 17$.
            Consequently~$\#G \geq 34$. Contradiction with minimality.

            \item
            Apparently~$b_1=b_2$ has fixed color~$0$ in~$B$.
            Hence, by Computation~\ref{comp:types}, $\#B \geq 15$.

@@

            The type of~$(a_1,a_2)$ in~$A$ cannot contain~$(1,0)$
            or~$(0,1)$ for otherwise~$G$ can be colored.
            It also cannot contain~$(1,1)$ as~$a_1$ and~$a_2$
            are adjacent.
            Thus both~$a_1$ and~$a_2$ have fixed color~$0$ in~$A$.
            Hence~$\#A\geq 17$ by Computation~\ref{comp:types}.
            Consequently~$\#G \geq 32$. Contradiction
            with minimality.
        \end{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 \neq b_2$.
            This leads to a contradication in the same way
            as in case~\ref{tricase1}.
    \item Apparently~$a_1 \neq a_2$ and~$b_1 \neq b_2$.
        The type of~$(a_1,a_2)$ in~$A$ cannot contain~$(0,0)$,
        for otherwise~$G$ is colorable.
        Similarly, the type of~$(b_1,b_2)$ in~$B$ cannot contain~$(0,0)$.
        Thus both~$\#A \geq 17$ and~$\#B \geq 17$.
        Hence~$\#G \geq 34$. Contradiction with minimality. \qedhere
    \end{enumerate}
\end{proof}
